---
title: "Machine Learning Project"
author: "Carlo Lopez, Vibha Naiknavare, Daniel Huang, George Cole "
date: "2023-09-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
library(ggplot2)
library(tidyverse)
```

Reading the Data
```{r}
# Reading data & taking out unwanted variables
spotify <- read.csv("spotify-2023.csv")
final_spotify_data <- spotify[,-c(3,10,11,12,13,14)]

# Cleaning data for summary (removing name, artist, and NA's)
spotify_summary_data <- final_spotify_data[,-c(1,2)]
spotify_summary_data$streams <- as.numeric(spotify$streams)
spotify_summary_data$key <- as.factor(spotify$key)
spotify_summary_data$mode <- as.factor(spotify$mode)

which(is.na(spotify_summary_data))

spotify_summary_data <- na.omit(spotify_summary_data)

# Summary Data 
summary(spotify_summary_data)

# Final test data
spotify_data <- spotify_summary_data[,-c(1,2,3,4,5)]
```

Variables: 
track_name: Name of the song
artist(s)_name: Name of the artist(s) of the song
artist_count: Number of artists contributing to the song
released_year: Year when the song was released
released_month: Month when the song was released
released_day: Day of the month when the song was released
in_spotify_playlists: Number of Spotify playlists the song is included in
in_spotify_charts: Presence and rank of the song on Spotify charts
streams: Total number of streams on Spotify
bpm: Beats per minute, a measure of song tempo
key: Key of the song
mode: Mode of the song (major or minor)
danceability_%: Percentage indicating how suitable the song is for dancing
valence_%: Positivity of the song's musical content
energy_%: Perceived energy level of the song
acousticness_%: Amount of acoustic sound in the song
instrumentalness_%: Amount of instrumental content in the song
liveness_%: Presence of live performance elements
speechiness_%: Amount of spoken words in the song

1 general linear regression analysis

3 individual analysis:
streams vs speechiness - line plot
streams vs danceability - line plot
streams vs keys - box-plot

correlation table
2 interaction terms analysis:
maybe keys + danceability or modes + speechniess

--> identify if the data is hugely impacted by extreme values

General Visualization of the data we have:
```{r}
options(scipen = 9)
Streams_density_vis <- ggplot(spotify_data, aes(x = streams)) + # Set X-axis as the number of streams
  geom_density(fill = "blue", alpha = 0.3) + # Use geom_density to get density plot
  theme_bw() + # Set theme for plot
  theme(panel.grid.major = element_blank(), # Turn of the background grid
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  labs(x = "Streams", # Set plot labels
       title = "Density plot of Streams")

Streams_density_vis # Generate plot
```
This is a pretty right skewed distribution. Most of the songs have their number of streams less than 1 billion, while some songs are extremely popular with more than 2 billion streams. We should apply a log-transform
```{r}
log_streams <- log(spotify_data$streams + 1)

summary(log_streams)

spotify_data$streams <- log_streams

spotify_data <- na.omit(spotify_data)

spotify_data <- spotify_data %>%
  rename("log_streams" = streams)

```


Regression Analysis: Which factors influence streams
```{r}
fit_1 <- lm(log_streams ~.,  # Set formula
            data = spotify_data) # Set dataset

summary(fit_1)
```
4 significant factors: danceability, acustincness, liveness, adn speechiness

Creating Visualizations based on significant variables

```{r}
spotify_vis <- ggplot(spotify_summary_data, 
              aes(y = streams, 
                  x = danceability_.)) + 
  geom_point(color = "blue", alpha = 0.3) + 
  geom_smooth(method = "lm") + 
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  labs(y = "Streams",
       x = "Danceability (%)",
       title = "Danceability on Streams")

spotify_vis
```



```{r}
spotify_vis2 <- ggplot(spotify_summary_data, 
              aes(y = streams, 
                  x = speechiness_. )) + 
  geom_point(color = "red", alpha = 0.4) + 
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  labs(y = "Streams",
       x = "Speechines",
       title = "Speechiness on Streams")

spotify_vis2
```

Packages:
```{r}
library(rpart)                      # Popular decision tree algorithm
library(rattle)                 # Fancy tree plot
library(rpart.plot)             # Enhanced tree plots
library(RColorBrewer)               # Color selection for fancy tree plot
library(party)                  # Alternative decision tree algorithm
library(partykit)               # Convert rpart object to BinaryTree
library(caret)  
library(reshape2) # Load reshape 2 for melting
library(DMwR) # Load data mining with R for SMOTE
library(splitstackshape) # Used for stratified sampling

```


Correlation and HeatMap
```{r}
numeric_columns <- spotify_data[, sapply(spotify_data, is.numeric)]

correlation_matrix <- cor(numeric_columns)
correlation_matrix

melted_corr_matrix <- melt(correlation_matrix)

ggplot(data = melted_corr_matrix, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0, limits = c(-1, 1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Correlation Heatmap")
```




Improved regression model Interaction terms 

```{r}


interaction_reg1 <- lm(log_streams ~ valence_.* danceability_.+.,  # Set formula
            data = spotify_data)

interaction_reg
summary(interaction_reg1)

interaction_terms <- ggplot(spotify_data, aes(y = interaction(valence_., danceability_.), color = valence_., x = danceability_.)) +
  geom_point(alpha = 0.3) +
  geom_smooth(aes(group = valence_.), method = "lm") +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  labs(y = "Interaction (valence_. & Danceability)",
       x = "Danceability",
       title = "Valence Relationship with Danceability")

print(interaction_terms)


colnames(spotify_data)
```




Decision Tree:
```{r}
options(scipen = 0)

tree_1 <- rpart(log_streams ~., # Set tree formula
data = spotify_data) # Set dataset
par(xpd = NA) # Set this avoid cut-off text
plot(tree_1)  # Plot tree
text(tree_1, digits = 3) # Add text

rpart.plot(tree_1, digits = 7, extra = 1)

```


Random Forest:
```{r}
library(randomForest)
library(rpart) # Load rpart for decision trees
library(caret) # Used for analysing results
library(splitstackshape) # Used for stratified sampling



set.seed(12)
length(spotify_data$log_streams)
# Sample all unique elements for train.rows2
train.rows2 <- sample(spotify_data$log_streams, 700, replace = FALSE)

# Identify the remaining elements for valid.rows2
remaining_elements <- setdiff(spotify_data$log_streams, train.rows2)

# Sample all remaining unique elements for valid.rows2
valid.rows2 <- sample(remaining_elements, length(remaining_elements), replace = FALSE)

train.df2 <- spotify_data[train.rows2,]

valid.df2 <- spotify_data[valid.rows2,]

tree_model <- rpart(log_streams ~., # Set tree formula
                data = train.df2) # Set dataset
tree_preds <- predict(tree_model, valid.df2)

# Calculate RMSE
rmse <- sqrt(mean((tree_preds - valid.df2$log_streams)^2))

# Print RMSE
print(paste("Root Mean Squared Error (RMSE):", round(rmse, 2)))

set.seed(258506) # Set random number generator seed for reproducability
# Use random forest to do bagging
bag_mod <- randomForest(log_streams ~., # Set tree formula
                data = train.df2, # Set dataset
                mtry = 10, # Set mtry to number of variables 
                ntree = 200) # Set number of trees to use
bag_mod # View model
```

XGBoost:
```{r}
library(xgboost)
set.seed(111111)

dtrain <- xgb.DMatrix(data = as.matrix(as.numeric(train.df2[,-which(names(train.df2) == "log_streams")])), label = train.df2$log_streams)

bst_1 <- xgboost(data = dtrain, # Set training data
               nrounds = 100, # Set number of rounds
               verbose = 1, # 1 - Prints out fit
               print_every_n = 20# Prints out result every 20th iteration
 ) 

Library(Metrics)

rmse(actual, predicted)
```

